---
title: "Dysregulation of Cerebral Hemoglobin Content in Malawian Children with Malaria"
author: "Rachel Smith"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, comment = "", fig.height = 8, fig.width = 10)

library(tidyverse)
library(janitor)
library(factoextra)
library(ggrepel)
library(ggpubr)
library(tidymodels)
library(vip)
library(scales)
library(reshape)


load("model1.Rdata")
load("model2.Rdata")

```

```{r data}

# read in data
df_hbox <- read.csv("Data/98 first visit records with cerebral oxygenation stats_trimmed.csv") %>% as_tibble()

# get rid of "Other"
df_hbox <- df_hbox %>% 
  filter(Status != "Other") %>% 
  mutate(Status = factor(Status, levels = c("HV", "UM", "CM")))

# identify duplicated rows (TM0003, TM2001)
df_hbox <- df_hbox %>% filter(duplicated(Subject.ID..NIAID.) == FALSE)

# Convert age to numeric
df_hbox <- df_hbox %>% 
  mutate(Age = gsub(" Years, ", "_", Age)) %>% 
  mutate(Age = gsub(" Months ", "", Age)) %>% 
  separate(Age, into = c("Years", "Months"), sep = "_") %>% 
  mutate(Years = as.numeric(Years)) %>% 
  mutate(Months = as.numeric(Months)/12) %>% 
  mutate(Age = Years + Months) %>% 
  dplyr::select(-c(Years, Months))

# Convert sex to factor
df_hbox <- df_hbox %>% mutate(Sex = as.factor(Sex))

# Trim for variables of interest (VOI)
df_hbox_trim <- df_hbox %>% 
  dplyr::select(c(Subject.ID..NIAID., Status, Glucose, Hematocrit, Lactate, Temperature,
                  Arginine.umol.L, Haptoglobin..mg.dl., Hemoglobin..uM., Whole.Blood.Nitrite, 
                  cerebral.hbdeox, cerebral.hbdeox.alpha, cerebral.hbox, cerebral.hbox.alpha,
                  cerebral.sat, cerebral.sat.alpha, cerebral.thc, cerebral.thc.alpha, cerebral.thc.norm.std)) %>% 
  
  dplyr::rename("subject_id" = "Subject.ID..NIAID.", "Arginine (umol/L)" = "Arginine.umol.L", "Haptoglobin (mg/dL)" = "Haptoglobin..mg.dl.", 
                "Hemoglobin (uM)" = "Hemoglobin..uM.", "WB_Nitrite" = "Whole.Blood.Nitrite", "DeoxyHb" = "cerebral.hbdeox", 
                "DeoxyHb_alpha" = "cerebral.hbdeox.alpha", "OxyHb" = "cerebral.hbox", "OxyHb_alpha" = "cerebral.hbox.alpha", 
                "O2sat" = "cerebral.sat", "O2sat_alpha" = "cerebral.sat.alpha", "THC" = "cerebral.thc", 
                "THC_alpha" = "cerebral.thc.alpha", "THC_sd" = "cerebral.thc.norm.std") 

# impute missing data with median by group
f_impute <- function(x, na.rm = TRUE) (replace(x, is.na(x), median(x, na.rm = na.rm)))

df_hbox_impute <- df_hbox_trim %>% 
  group_by(Status) %>% 
  mutate_at(3:ncol(.), f_impute) %>% 
  ungroup()

# pivot long
df_hbox_long <- df_hbox_trim %>% pivot_longer(cols = 3:ncol(.), names_to = "measurement", values_to = "value")

```

### Table 1: Summary of hematological data by patient group

```{r table1}

# number of subjects 
df_status <- df_hbox_impute %>% 
  count(Status) %>% 
  pivot_wider(1:2, names_from = "Status", values_from = "n") %>% 
  mutate(variable = "Number of subjects") %>% 
  dplyr::select(variable, everything()) %>% 
  mutate_if(is.numeric, as.character)

# Sex
df_sex <- df_hbox %>% 
  group_by(Status) %>% 
  count(Sex) %>% 
  filter(Sex != "") %>% 
  pivot_wider(names_from = "Sex", values_from = "n") %>% 
  mutate(perc_fem = round(Female/(Female + Male)*100, 2)) %>% 
  
  pivot_wider(c(1,4), names_from = "Status", values_from = "perc_fem") %>% 
  mutate(variable = "Sex (% female)") %>% 
  dplyr::select(variable, everything()) %>% 
  mutate_if(is.numeric, as.character)

# Age, hematocrit, and lactate
df_hbox %>% 
  group_by(Status) %>% 
  summarise(`Age (years)` = median(Age, na.rm = TRUE), age_range = range(Age, na.rm = TRUE), 
            `Hematocrit (%)` = median(Hematocrit, na.rm = TRUE), hem_range = range(Hematocrit, na.rm = TRUE),
            `Lactate (mmol/L)` = median(Lactate, na.rm = TRUE), lactate_range = range(Lactate, na.rm = TRUE),
            `Arginine (umol/L)` = median(Arginine.umol.L, na.rm = TRUE), arg_range = range(Arginine.umol.L, na.rm = TRUE),
            `Temperature (°C)` = median(Temperature, na.rm = TRUE), temp_range = range(Temperature, na.rm = TRUE)) %>% 
  mutate_if(is.numeric, round, 1) %>% 
  
  mutate(`Age (years)` = paste0(`Age (years)`, " (", age_range[1], ", ", age_range[2], ")"),
         `Hematocrit (%)` = paste0(`Hematocrit (%)`, " (", hem_range[1], ", ", hem_range[2], ")"),
         `Lactate (mmol/L)` = paste0(`Lactate (mmol/L)`, " (", lactate_range[1], ", ", lactate_range[2], ")"),
         `Arginine (umol/L)` = paste0(`Arginine (umol/L)`, " (", arg_range[1], ", ", arg_range[2], ")"),
         `Temperature (°C)` = paste0(`Temperature (°C)`, " (", temp_range[1], ", ", temp_range[2], ")")) %>% 
  dplyr::select(-c(age_range, hem_range, lactate_range, arg_range, temp_range)) %>% 
  unique() %>% 
  t() %>% 
  row_to_names(row_number = 1) %>% 
  as_tibble(rownames = "variable") %>% 
  add_row(df_status, .before = 1) %>% 
  add_row(df_sex, .before = 2) %>% 
  
  dplyr::rename(" " = "variable",
                "Healthy volunteers (HV)" = "HV",
                "Uncomplicated malaria patients (UM)" = "UM",
                "Cerebral malaria patients (CM)" = "CM") %>% 
  knitr::kable(align = rep('c', 4))

```
**Table 1.** A summary of subject information and hematological data. Group patient data presented as median (range). Note: sex not recorded for 3 CM patients.

### Figure 1: Visual summary of NIRS variables

```{r fig1}

df_hbox_voi <- df_hbox_impute %>% 
  dplyr::select(c(Status, DeoxyHb, DeoxyHb_alpha, OxyHb, OxyHb_alpha, O2sat, O2sat_alpha, THC, THC_alpha, THC_sd)) %>% 
  dplyr::filter(DeoxyHb_alpha < 3) %>% 
  dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha", "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", 
                "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha", "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha")

df_hbox_voi_long <- df_hbox_voi %>% 
  pivot_longer(2:10, names_to = "variable", values_to = "value") %>%
  mutate(variable = factor(variable, levels = c("Deoxy Hb", "Oxy Hb", "O2 sat", "THC", "THC sd", 
                                                "Deoxy Hb α", "Oxy Hb α", "O2 sat α", "THCα"))) 

my_col <- c("CM" = "#00BFC4", "HV" = "#7CAE00", "UM" = "#F8766D")
  
df_hbox_voi_long %>% 
  
  ggplot(aes(x = Status, y = value)) +
  geom_violin(aes(color = Status)) +
  geom_jitter(position = position_jitter(0.2), shape = 1) +
  stat_summary(fun = "median", geom = "crossbar", aes(color = Status), size = 0.2, width = 0.5) +
  facet_wrap(~variable, scales = "free_y") +
  
  scale_color_manual(values = my_col) +
  theme_bw() +
  theme(legend.position = "none")  +
  
  theme(strip.text = element_text(size = 15),
        axis.title = element_text(size = 15),
        axis.title.y = element_blank(),
        axis.text = element_text(size = 15))

```
**Figure 1.** Differences in variables measured by Oxiplex and associated alpha values among patient groups. Note: removed one outlier from DeoxyHb alpha (3.52)

### Figure 2: Random forest modeling using all outcomes and predictors

#### A. PCA
```{r fig2a}

# convert tibble to df
df_hbox_impute_df <- df_hbox_impute %>% 
  dplyr::select(-c(subject_id, Status)) %>% 
  dplyr::rename("Whole blood nitrite" = "WB_Nitrite", "THC sd" = "THC_sd", "THCα" = "THC_alpha", 
                "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha",
                "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha",
                "Hemoglobin" = "Hemoglobin (uM)", "Haptoglobin" = "Haptoglobin (mg/dL)", "Arginine" = "Arginine (umol/L)") %>% 
  as.data.frame(df_hbox_impute) 
rownames(df_hbox_impute_df) <- df_hbox_impute$subject_id

# compute PCA
my_pca1 <- prcomp(df_hbox_impute_df, scale = TRUE)

# define groups
groups1 <- as.factor(df_hbox_impute$Status)

# Eigenvalues
pca_eig_val1 <- get_eigenvalue(my_pca1)

# extract variables
df_vars1 <- my_pca1$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca1$rotation))

# extract patients
df_pnts1 <- my_pca1$x %>% as_tibble() %>% mutate(group = groups1)

# plot

  # plot patients
  ggplot() +
  geom_point(data = df_pnts1, 
             mapping = aes(x = PC1, y = PC2, color = group), 
             size = 3) +
  stat_conf_ellipse(data = df_pnts1, 
                    mapping = aes(x = PC1, y = PC2, color = group, fill = group),
                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(data = df_pnts1, 
            mapping = aes(x = PC1, y = PC2, color = group, shape = group),
            na.rm = FALSE) +
  
  # plot variables
  geom_point(data = df_vars1, aes(x = PC1*10, y = PC2*10)) +
  geom_text_repel(data = df_vars1, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
  geom_segment(data = df_vars1, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +

  # draw lines
  geom_vline(xintercept = 0, lty = 2, color = "black") +
  geom_hline(yintercept = 0, lty = 2, color = "black") +
  
  # design
  scale_color_manual(values = my_col) +
  scale_fill_manual(values = my_col) +
  xlab(paste0("PC1: ", round(pca_eig_val1$variance.percent[1], 2), "% of variance")) +
  ylab(paste0("PC2: ", round(pca_eig_val1$variance.percent[2], 2), "% of variance")) +

  theme_bw() +
  theme(axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))

```

#### B. Confusion matrix
```{r fig2b_prep, eval = FALSE}

# Data
df_hbox_model1 <- df_hbox_trim %>% 
  dplyr::select(-subject_id) %>% 
  dplyr::rename("Whole blood nitrite" = "WB_Nitrite", "THC sd" = "THC_sd", "THCα" = "THC_alpha", 
                "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha",
                "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha",
                "Hemoglobin" = "Hemoglobin (uM)", "Haptoglobin" = "Haptoglobin (mg/dL)", "Arginine" = "Arginine (umol/L)")

# Build a model

  # resamples (not a lot of data, need to use cross validation)
  set.seed(123)
  hbox_boot1 <- bootstraps(df_hbox_model1, strata = Status)

  # Create recipe
  ranger_recipe <- 
    recipe(formula = Status ~ ., data = df_hbox_model1) %>% 
    
    # BoxCox transformation to normality
    step_BoxCox(all_predictors(), -all_outcomes()) %>% 
    
    # remove variables with non-zero variance
    step_nzv(all_predictors(), -all_outcomes()) %>% 
    
    # impute missing data using K-nearest neighbors
    step_knnimpute(all_predictors(), -all_outcomes()) 
  
  # Random forest model specifications (set for tuning)
  ranger_spec <- 
    rand_forest(mtry = tune(), min_n = tune(), trees = 1000) %>% 
    set_mode("classification") %>% 
    set_engine("ranger") 
  
  # Create workflow
  ranger_workflow <- 
    workflow() %>% 
    add_recipe(ranger_recipe) %>% 
    add_model(ranger_spec) 

# Tune model

  # Initial tune on bootstraps
  set.seed(84374)
  ranger_tune <-
    tune_grid(ranger_workflow, 
              resamples = hbox_boot1, 
              grid = 11)
  
# Finalize model

  final_rf <- ranger_workflow %>% 
    finalize_workflow(select_best(ranger_tune, metric = "accuracy")) 
  
  final_rs_res <- final_rf %>% 
    fit_resamples(resamples = hbox_boot1,
                  metrics = metric_set(roc_auc, accuracy, sens, spec),
                  control = control_resamples(save_pred = TRUE))

  
# save objects so model doesn't rerun every time you knit
  save(df_hbox_model1, hbox_boot1, ranger_recipe, ranger_spec, ranger_tune, final_rs_res, file = "model1.Rdata")

```

```{r fig2b_plot}

final_rs_res %>% 
  conf_mat_resampled() %>%
  
  # plot
  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
  geom_tile() +
  geom_text(size = 8) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
  
  theme_bw() +
  theme(axis.title = element_text(size = 25),
        axis.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 15))

```

#### C. Variable importance
```{r fig2c}

# new specification for ranger model
imp_spec <- ranger_spec %>% 
  finalize_model(select_best(ranger_tune, metric = "accuracy")) %>% 
  set_engine("ranger", importance = "permutation")

# plot importance 
workflow() %>% 
  add_recipe(ranger_recipe) %>% 
  add_model(imp_spec) %>% 
  fit(df_hbox_model1) %>% 
  pull_workflow_fit() %>% 
  vip(aesthetics = list(alpha = 0.8, fill = "midnightblue")) +
  
  theme_bw() +
  theme(axis.text = element_text(size = 20),
        axis.title = element_text(size = 25))

```

**Figure 2.** Random forest modeling demonstrates a high degree of discriminatory power between healthy patients and patients with malaria (uncomplicated and cerebral) using primarily on hematological parameters. **(A)** Principal component analysis (PCA) of study patients using hematological and NIRS data. Variables are plotted as vectors in the direction and magnitude of the PC to which they contribute. Color indicates patient status, ellipses show median and 95% confidence interval on the cluster. **(B)** Confusion matrix representing the results of random forest modeling averaged across 25 bootstrap resamples. Freq indicates the average number of times that the event (shown by the x- and y-axes) occurred. **(C)** Relative contribution of the variables to the discriminatory power of the model (showing top 10 only).

### Figure 3: Logistic regression on CM vs UM

#### A. PCA

```{r fig3a}
# select variables of interest
df_hbox_log_voi <- df_hbox_impute %>% 
  filter(Status != "HV") %>% 
  dplyr::select(2, 11:19)

# convert tibble to df
df_hbox_voi_df <- df_hbox_log_voi %>% 
  dplyr::select(-Status) %>% 
  dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha", "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", 
                "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha", "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha") %>% 
  as.data.frame() 
rownames(df_hbox_voi_df) <- df_hbox_impute %>% dplyr::filter(Status != "HV") %>% .$subject_id

# compute PCA
my_pca2 <- prcomp(df_hbox_voi_df, scale = TRUE)

# define groups
groups2 <- as.factor(df_hbox_log_voi$Status)

# Eigenvalues
pca_eig_val2 <- get_eigenvalue(my_pca2)

# two data frames for plotting
df_vars2 <- my_pca2$rotation %>% as_tibble() %>% mutate(variable = rownames(my_pca2$rotation))
df_pnts2 <- my_pca2$x %>% as_tibble() %>% mutate(group = groups2) 

# plot

  # plot patients
  ggplot() +
  geom_point(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group), 
             size = 3) +
  stat_conf_ellipse(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group, fill = group),
                    level = 0.95, npoint = 100, bary = TRUE, alpha = 0.1, geom = "polygon") +
  stat_mean(data = df_pnts2, mapping = aes(x = PC1, y = PC2, color = group, shape = group),
            na.rm = FALSE) +
  
  # plot variables
  geom_point(data = df_vars2, aes(x = PC1*10, y = PC2*10)) +
  geom_text_repel(data = df_vars2, aes(x = PC1*10, y = PC2*10, label = variable, size = 10), show.legend = FALSE) +
  geom_segment(data = df_vars2, aes(x = 0, y = 0, xend = PC1*10, yend = PC2*10), arrow = arrow()) +

  # draw lines
  geom_vline(xintercept = 0, lty = 2, color = "black") +
  geom_hline(yintercept = 0, lty = 2, color = "black") +
  
  # design
  xlab(paste0("PC1: ", round(pca_eig_val2$variance.percent[1], 2), "% of variance")) +
  ylab(paste0("PC2: ", round(pca_eig_val2$variance.percent[2], 2), "% of variance")) +

  theme_bw() +
  theme(axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))

```

#### B. Confusion matrix

```{r fig3b_prep, eval = FALSE}

# convert to 0s and 1s
df_hbox_voi_01 <- df_hbox_log_voi %>% mutate(Status = ifelse(Status == "CM", 1, 0) %>% as.factor())

# select only significant variables
df_hbox_model2 <- df_hbox_voi_01 %>% 
  dplyr::select(c(Status, THC_alpha, THC_sd)) %>% 
  dplyr::rename("THCα" = "THC_alpha", "THC sd" = "THC_sd") %>% 
  mutate(`1/10 * THCα` = THCα/10) %>% 
  dplyr::select(-THCα)

# resample
set.seed(789)
hbox_boot2 <- bootstraps(df_hbox_model2, times = 1000)

# recipe
hbox_rec2 <- recipe(Status ~ ., data = df_hbox_model2) %>% 
  step_BoxCox(all_predictors()) %>% 
  step_nzv(all_predictors())

# workflow
hbox_wf2 <- workflow() %>% 
  add_recipe(hbox_rec2)

# logistic regression model spec
glm_spec2 <- logistic_reg() %>% 
  set_engine("glm")

glm_res2 <- hbox_wf2 %>% 
  add_model(glm_spec2) %>%
  fit_resamples(
    resamples = hbox_boot2,
    metrics = metric_set(roc_auc, accuracy, sensitivity, specificity),
    control = tune::control_resamples(save_pred = TRUE, verbose = TRUE)
  )

# select best model to be used for evaluation
glm_best2 <- glm_res2 %>% select_best("roc_auc")

# finalize model
hbox_final2 <- hbox_wf2 %>% 
  add_model(glm_spec2) %>% 
  finalize_workflow(glm_best2) %>%
  fit(df_hbox_model2) %>%
  pull_workflow_fit()

# save objects so model doesn't rerun 
save(glm_res2, hbox_final2, file = "model2.Rdata")

```   

```{r fig3b_plot}

glm_res2 %>% 
  conf_mat_resampled() %>% 
  mutate(Prediction = ifelse(Prediction == 1, "CM", "UM")) %>% 
  mutate(Truth = ifelse(Truth == 1, "CM", "UM")) %>% 
  
  # plot
  ggplot(aes(x = Truth, y = Prediction, fill = Freq, label = Freq)) +
  geom_tile() +
  geom_text(size = 6) +
  scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
  
  theme_bw() +
  theme(axis.title = element_text(size = 25),
        axis.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.text = element_text(size = 15))

```

**Figure 3** Logistic regression model discriminating CM vs UM patients using only THCα and THC sd as predictors. **(A)** PCA of patients with CM and UM as determined by NIRS variables. Color indicates patient status, ellipses show median and 95% confidence interval on the cluster. **(B)** Confusion matrix representing the results of the logistic regression model averaged across 1000 bootstrap resamples. Freq indicates the average number of times that the event (shown by the x- and y-axes) occurred.

### Table 2. Logistic regression model evaluation

#### A. Model metrics
```{r table2a}

collect_metrics(glm_res2) %>% 
  dplyr::select(.metric, mean, std_err) %>% 
  mutate(mean = mean*100, std_err = std_err*100) %>% 
  mutate(Metric = c("Accuracy", "ROC AUC", "Sensitivity", "Specificity")) %>% 
  dplyr::select(c("Metric", "mean", "std_err")) %>% 
  
  dplyr::rename("Mean across resamples" = "mean", "Standard error" = "std_err") %>% 
  mutate_if(is.numeric, round, 2) %>% 
  as.data.frame() %>%
    
  knitr::kable(align = rep('c', 4))

```

#### B. Coefficients
```{r table2b}

hbox_final2 %>% 
  tidy() %>% 
  mutate(p.value = as.character(p.value)) %>% 
  mutate_if(is.numeric, round, 2) %>% 
  mutate(p.value = as.numeric(p.value)) %>% 
  mutate(p.value = scientific(p.value, digits = 3)) %>% 
  dplyr::rename("Term" = "term", "Coefficient" = "estimate", "Standard error" = "std.error", 
                "Statistic" = "statistic", "p-value" = "p.value") %>% 
  
  as.data.frame() %>%
  knitr::kable(align = rep('c', 5))


```

**Table 2.** Evaluation of the logistic regression model across 1000 bootstrap resamples. 

### Figure 4: Relationship between THCα and THC sd

```{r fig4}

cm_model <- lm(THC_sd ~ THC_alpha, data = df_hbox_log_voi %>% filter(Status == "CM"))
um_model <- lm(THC_sd ~ THC_alpha, data = df_hbox_log_voi %>% filter(Status == "UM"))

df_hbox_log_voi %>% dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha") %>%
  
  ggplot(aes(x = THCα, y = `THC sd`, color = Status, fill = Status)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  
  geom_text(label = paste0("y = ", 
                           round(cm_model$coefficients[[1]], 2), 
                           " + ", 
                           round(cm_model$coefficients[[2]], 2), 
                           "x ; R^2 = ", 
                           round(summary(cm_model)$adj.r.squared, 2)),
            x = 1.3, y = 0.55,
            color = "#00BFC4",
            size = 6) +
  
    geom_text(label = paste0("y = ", 
                           round(um_model$coefficients[[1]], 2), 
                           " + ", 
                           round(um_model$coefficients[[2]], 2), 
                           "x ; R^2 = ", 
                           round(summary(um_model)$adj.r.squared, 2)),
            x = 1.3, y = 0.6,
            color = "#F8766D",
            size = 6) +

  theme_bw() +
  theme(axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20))

```

**Figure 4.** Scatter plot demonstrating the relationship between THCα and THC sd.

### Supplementary figure 1: Variable correlations 
```{r fig_s1, fig.show = "hold", out.width = "50%", fig.height = 12, fig.width = 15}

# calculate spearman correlations in each group

f_calc_cor <- function(df, status) {
  
  df %>%
    filter(Status == status) %>% 
    dplyr::select(-Status) %>% 
    cor(method = "spearman")
  
}

df_hbox_cor <- df_hbox_impute %>% 
  dplyr::rename("THC sd" = "THC_sd", "THCα" = "THC_alpha", "Deoxy Hb" = "DeoxyHb", "Oxy Hb" = "OxyHb", 
                "O2 sat" = "O2sat", "O2 sat α" = "O2sat_alpha", "Deoxy Hb α" = "DeoxyHb_alpha", "Oxy Hb α" = "OxyHb_alpha") %>% 
  dplyr::select(Status, `Deoxy Hb`, `Oxy Hb`, `O2 sat`, THC, `THC sd`, `Deoxy Hb α`, `Oxy Hb α`, `O2 sat α`, THCα)

m_cor_hv <- df_hbox_cor %>% f_calc_cor(status = "HV")

m_cor_um <- df_hbox_cor %>% f_calc_cor(status = "UM")

m_cor_cm <- df_hbox_cor %>% f_calc_cor(status = "CM")

# correlation matrix

# write plot function
f_plot_cor <- function(mat) {
  
  melt(mat) %>% 
    
    mutate_if(is.factor, factor, levels = c("Deoxy Hb", "Oxy Hb", "O2 sat", "THC", "THC sd", "Deoxy Hb α", "Oxy Hb α", "O2 sat α", "THCα")) %>% 
    
    ggplot(aes(x = X1, y = X2, fill = value, label = round(value, 2))) + 
    geom_tile() +
    scale_fill_gradient2(low = "#4575b4", high = "#d73027") +
    geom_text(size = 9) +
    
    xlab("") +
    ylab("") +
    theme(axis.text.x = element_text(angle = 45, hjust = 0.9),
          axis.text = element_text(size = 35),
          legend.text = element_text(size = 40),
          legend.title = element_text(size = 50),
          legend.key.size = unit(2, 'cm'),
          plot.title = element_text(size = 30)
    )
  
}

  # extract legend
  p_leg <- m_cor_hv %>% f_plot_cor() %>% get_legend()
  
  # create plots
  m_cor_hv %>% f_plot_cor() + ggtitle("HV variable correlations") + theme(legend.position = "none")
  
  m_cor_um %>% f_plot_cor() + ggtitle("UM variable correlations") + theme(legend.position = "none")
  
  m_cor_cm %>% f_plot_cor() + ggtitle("CM variable correlations") + theme(legend.position = "none")
  
  as_ggplot(p_leg)

```

**Figure S1.** Shown are the spearman correlations among variables measured by Oxiplex and their associated alpha values. Red indicates a positive correlation; blue indicates a negative correlation.

### Supplementary figure 2: Logistic regression curves of NIRS variables
```{r fig_s2}

df_hbox_voi_long %>% mutate(Status = as.numeric(ifelse(Status == "CM", 1, 0))) %>% 
  
  ggplot(aes(x = value, y = Status)) +
  geom_point() +
  geom_smooth(method = "glm", method.args = list(family = "binomial")) +
  facet_wrap(~variable, scales = "free_x") +
  
  ylab("Status: 1 = CM, 0 = UM") +
  xlab("") +
  
  theme_bw() +
  theme(axis.text = element_text(size = 20),
        axis.title = element_text(size = 25),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 20),
        strip.text = element_text(size = 20))

```

**Figure S2.** Plotted are the individual logistic regression curves of individual variables tested for their discriminatory power between UM and CM. On the y-axis, 1 represents CM while 0 represents UM. A steeper curve indicates higher disciminatory power